\lecture{15}{mon 18 oct 10:30}{Taxes}
\section{Elasticity \& Taxes}
With taxes into consideration, whichever supply curve is more elastic will endure less of the burden. The more inelastic one side is, the more of the burden they will have to bear. With taxes, you have to consider who submits the tax, and add it to that curve respectively. In most taxes, you will be applying them to the supply curve. 

I will be honest, I am super tired as I am writing this, so I do recommend checking the Elasticity 20.21 v5 slideshow posted in the unit guide. At the end of the slideshow it has the graphs and practice which I do recommend doing. Make sure to realize how TS, CS, and PS are being affected by the elasticity. 

\textbf{Summary}
\begin{itemize}
    \item Elasticity measures the responsiveness of Q to one of its determinants
    \item Price Elasticity of Demand equals percentage change in Q divided by percentage change of P
        \begin{itemize}
            \item When it's less than one, demand is inelastic. When its greater than one, demand is elastic
            \item When demand is inelastic, total revenue rises when prices rise
            \item When demand is elastic, total revenue falls when prices rise
            \item Same can be said for supply, just replace word demand for supply
        \end{itemize}
    \item Income Elasticity
        \begin{itemize}
            \item Measures how much quantity demanded responds to changes in buyers' incomes. 
            \item $E_Y < 0$ = inferior good
            \item  $0 < E_Y < 1$ = normal good-necessity
            \item  $E_Y > 1$ = normal good-luxury
        \end{itemize}
    \item Cross-Price Elasticity
        \begin{itemize}
            \item Measures how much demand for one good responds to changes in the price of another good.
            \item $E_{CP} < 0$ = complements
            \item  $E_{CP} = 0$ = unrelated
            \item $E_{CP} > 0$ = substitutes
        \end{itemize}
\end{itemize}

This is all you will need to know for the Unit 2 test, and make sure to review with the workbook and other resources such as the slides. 
